Algebraic Principles for Concurrency Verification Tools

### 2013-10-03 Paper/paper.tex
Started on language model section

@@

\tikzstyle{elem} = [circle]
\tikzstyle{line} = [draw,thick, -latex']

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Kleene Algebra}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rely/Guarantee}

$r, g \vdash \{p\}x\{q\} \longleftrightarrow p(r\|x)\le q \land x \le g$

\begin{prooftree}
\LeftLabel{Weakening}
\AxiomC{$r' \le r$}
\AxiomC{$g \le g'$}
\AxiomC{$p \le p'$}
\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
\AxiomC{$q' \le q$}
\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
\end{prooftree}

\begin{prooftree}
\LeftLabel{Sequential}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{q\}y\{s\}$}
\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
\end{prooftree}

\begin{prooftree}
\LeftLabel{Parallel}
\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

\begin{prooftree}
\LeftLabel{Choice}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{p\}y\{q\}$}
\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
\end{prooftree}

\newpage
\section{Algebra for Rely/Guarantee}

@@

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's \emph{Nitpick} counterexample generator
to construct models which violate each of these axioms, yet satisfy
all others. The models thus constructed for (\ref{rg1}) and
(\ref{rg3}) are shown in figures \ref{fig:rg1} and \ref{fig:rg3}
respectively. The model constructed for (\ref{rg2}) is just the two
element rely/guarantee algebra with $0$ and $1$.

@@

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
models. This is because any finite $K$ is a weak left quantale with
respect to both sequential and parallel composition. In a quantale,
fixpoint fusion laws can be used to prove that
\begin{align*}
r\|x^+ = r\|(\mu y.\; x + xy) = \mu y.\; r\|x + r\|xy = (r\|x)^+.
\end{align*}
Thus it is impossible to construct a model which demonstrates
that (\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Language Model}

>>>>

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}

\newcommand{\udagger}{\text{$\rotatebox[origin=c]{180}{$\dagger$}$}}

@@

\tikzstyle{elem} = [circle]
\tikzstyle{line} = [draw,thick, -latex']
\tikzstyle{rel} = [draw,thin,dashed, -latex']

@@

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

x

related work?

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Kleene Algebra}
\label{sec:KA}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebra for Rely/Guarantee}
\label{sec:RG}

@@

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's
\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample generator
to construct models which violate each of these axioms, yet satisfy
all others. The models thus constructed for (\ref{rg1}) and
(\ref{rg3}) are shown in Figures \ref{fig:rg1} and \ref{fig:rg3}
respectively. The model constructed for (\ref{rg2}) is just the two
element rely/guarantee algebra with $0$ and $1$.

@@

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
models. This is because any finite $K$ is a weak left quantale with
respect to both sequential and parallel composition. In a quantale,
fixpoint fusion and induction laws can be used to prove that
\begin{align*}
r\|x^+ = r\|(\mu y.\; x + xy) = \mu y.\; r\|x + r\|xy \le (r\|x)^+.
\end{align*}
Thus it is impossible to construct a finite model demonstrating
that (\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).

@@

Jones quintuples can be encodeded in this setting as
\begin{align}
r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g.
\end{align}
Using this encoding we can derive all the standard rules for
rely/guarantee, shown in Figure \ref{fig:rgrules}.

We can expand our rely/guarantee algebras to have residuals
axiomatised by Galois connections for both parallel and sequential
composition, i.e.
\begin{align}
xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
x\|y \le z &\iff y \le x/z \label{res3}.
\end{align}
For sequential composition, (\ref{res1}) and (\ref{res2}) essentially
provide a link between weakest-precondition style semantics and hoare
logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
understood similarly as follows; it is the weakest program such that
when placed in parallel with $x$, the composition behaves as $z$.

By using residuals quintuples can be encoded in the following
equivalent way,
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program
refinement, as $r/(p \rightarrow q) \sqcap g$ defines the weakest
program that when placed in parallel with interference from $r$, and
guaranteeing interference at most $g$, goes from $p$ to $q$.

\begin{figure}[tbh]
\centering
\begin{prooftree}
\RightLabel{Skip}
\AxiomC{$pr \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Weakening}
\AxiomC{$r' \le r$}
\AxiomC{$g \le g'$}
\AxiomC{$p \le p'$}
\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
\AxiomC{$q' \le q$}
\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Sequential}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{q\}y\{s\}$}
\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Parallel}
\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Choice}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{p\}y\{q\}$}
\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
\end{prooftree}

\begin{prooftree}
\RightLabel{Star}
\AxiomC{$pr \le p$}
\AxiomC{$r, g \vdash \{p\}x\{p\}$}
\BinaryInfC{$r, g \vdash \{p\}x^\star\{p\}$}
\end{prooftree}
\caption{Rely/guarantee inference rules}
\label{fig:rgrules}
\end{figure}
x

x

x

x

x

x

x

x

x

x
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Concurrent Kleene Algebra with Tests}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finite Language Model}

In this section, we construct a finite language model satisfying the
axioms in Section \ref{sec:RG}. Restricting our attention to finite
languages means that we do not need to concern ourselves with
termination side-conditions.

The shuffle product $\sha$ on a pair of finite words is defined inductively as
\begin{align*}
\epsilon \sha s &= \{s\},\\
s \sha \epsilon &= \{s\},\\
as \sha bt &= a(s \sha bt) \cup b(as \sha t).
\end{align*}
Shuffle is associative, commutative, and distributes over arbitrary
joins.

\begin{lem}
$(\mathcal{P}((\Sigma\times\Sigma)^*), \cup, \cdot, \sha, \emptyset, \{\epsilon\})$ is a weak trioid.
\end{lem}

Our alphabet comprises relations ($R,S,T,\dots$) over a state space
$\Sigma$. An example of a word in
$(\{\sigma_1,\sigma_2,\sigma_3\}^2)^*$ is shown below. This word only
has a single valid \emph{trace}, which is
$\sigma_1\rightarrow\sigma_1\rightarrow\sigma_2\rightarrow\sigma_3\rightarrow\sigma_2$.
We define an operator $\tr(X)$ which returns all the traces in a language
$X$.

\begin{center}
\begin{tikzpicture}[x=1cm,auto]
  \node (center) {};
  \node [elem] (s12) {$\sigma_2$};
  \node [elem, above of=s12] (s11) {$\sigma_1$};
  \node [elem, below of=s12] (s13) {$\sigma_3$};
  \node [right of=s13, node distance=0.75cm] (r1) {};
  \node [below of=r1, node distance=0.5cm] (t4) {R};

  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
  \node [elem, above of=s22] (s21) {$\sigma_1$};
  \node [elem, below of=s22] (s23) {$\sigma_3$};
  \node [right of=s23, node distance=0.75cm] (r2) {};
  \node [below of=r2, node distance=0.5cm] (t4) {S};

  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
  \node [elem, above of=s32] (s31) {$\sigma_1$};
  \node [elem, below of=s32] (s33) {$\sigma_3$};
  \node [right of=s33, node distance=0.75cm] (r3) {};
  \node [below of=r3, node distance=0.5cm] (t4) {T};

  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
  \node [elem, above of=s42] (s41) {$\sigma_1$};
  \node [elem, below of=s42] (s43) {$\sigma_3$};
  \node [right of=s43, node distance=0.75cm] (r4) {};
  \node [below of=r4, node distance=0.5cm] (t4) {U};

  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
  \node [elem, above of=s52] (s51) {$\sigma_1$};
  \node [elem, below of=s52] (s53) {$\sigma_3$};

  \path [line] (s11) -- (s21);
  \path [line] (s21) -- (s32);
  \path [line] (s32) -- (s43);
  \path [line] (s43) -- (s52);

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
  \path [rel] (s32) -- (s42);
  \path [rel] (s42) -- (s51);
  \path [rel] (s41) -- (s52);
\end{tikzpicture}
\end{center}

Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
we inductively generate the stutter/mumble language $w^\dagger$ for a
word $w$ in a language over $\Sigma^2$ as follows. Assume $R,S,T \in
\Sigma^2$ and $u,v,w \in (\Sigma^2)^*$. First, $Iw \in
w^\dagger$. Secondly, if $uRv \in x^\dagger$ then $uIRv \in
x^\dagger$ and $uRIv \in x^\dagger$ (\emph{stuttering}). Thirdly, if
$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$
(\emph{mumbling}). The stutter/mumble closure of a language is
defined by lifting this operation to languages in the obvious way.

Next, we define the following language for a word $w = R_0R_1\dots
R_n$. The language $w^\udagger$ contains all strings $S_0S_1\dots S_n$
such that for all $m \le n$, $S_m \subseteq R_m$. Again this operation
is lifted to languages in the same way as before. Both these closure
operators are combined then into a single closure operator,
$X^\ddagger$, which satisfies the properties
\begin{align*}
(X^\ddagger \sha Y)^\ddagger &= (X \sha Y)^\ddagger,\\
(X^\ddagger Y)^\ddagger = (&XY)^\ddagger = (XY^\ddagger)^\ddagger,\\
(X \cup Y)^\ddagger &= (X^\ddagger \cup Y^\ddagger),\\
(X^\ddagger \cap Y^\ddagger)^\ddagger &= (X^\ddagger \cap Y^\ddagger),\\
\tr(X^\ddagger) &\subseteq X^\ddagger.
\end{align*}
For all binary operators $\bullet$, we can therefore define
$\bullet_\ddagger$ as $\lambda X Y. (X^\ddagger\bullet
Y^\ddagger)^\ddagger$ and show that
$(\mathcal{P}((\Sigma^2)^*)^\ddagger,\cup_\ddagger,\cdot_\ddagger,\sha_\ddagger,\emptyset,{\epsilon}^\ddagger)$
is a weak trioid.
